---
title: "JForster_Assignment3"
author: 'jfor: Josh Forster'
date: "2023-02-08"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
library(pracma)
```

(1) What is the rank of the matrix A?
The rank of matrix A = 4 \[ r = 4 \] as the reduced row echelon form is the identity matrix
```{r}
one_mat = matrix(c(1,-1,0,5,2,0,1,4,3,1,-2,-2,4,3,1,-3),ncol=4,nrow=4)
print(pracma::rref(one_mat))
pracma::Rank(one_mat)
``` 


(2) Given an mxn matrix where m > n, what can be the maximum rank? The minimum rank, assuming that the matrix is non-zero?

The maximum rank of this m x n matrix is n if all columns are pivot columns and the minimum rank is 1

3) What is the rank of matrix B?

The rank of matrix B is 1 \[ r = 1 \]
```{r}
three_mat = matrix(c(1,3,2,2,6,4,1,3,2),ncol=3,nrow=3)
pracma::rref(three_mat)
pracma::Rank(three_mat)
```

3. Compute the eigenvalues and eigenvectors of the matrix A. You’ll need to show your
work. You’ll need to write out the characteristic polynomial and show your solution.

$$X = \begin{bmatrix}1 & 2 & 3 \\
0 & 4 & 5 \\
0 & 0 & 6
\end{bmatrix}$$


$$X = \begin{bmatrix}1-\lambda & 2  & 3 \\
0 & 4 - \lambda & 5 \\
0 & 0 & 6 - \lambda
\end{bmatrix}$$

\[det(A-\lambda I_{n}) = 0 \]

\[(1-\lambda)((4-\lambda)(6-\lambda) - 5*0))-2(0(6-\lambda)-5*0) + 3(0*0 - 0(4-\lambda)=0) \]

\[(1-\lambda)(4-\lambda)(6-\lambda) - 2*(0)+3*0) = 0\]

\[\lambda = {1,6,4} \]

All algebraic multiplicities evaluate to 1


\[\epsilon_{\lambda=1}\]

$$X = \begin{bmatrix}1-1 & 2  & 3 \\
0 & 4 - 1 & 5 \\
0 & 0 & 6 - 1
\end{bmatrix}$$

$$X = \begin{bmatrix}0 & 2  & 3 \\
0 & 3 & 5 \\
0 & 0 & 5
\end{bmatrix}$$


$$X = \begin{bmatrix}1-1 & 2  & 3 \\
0 & 4 - 1 & 5 \\
0 & 0 & 6 - 1
\end{bmatrix}$$

Row Reduction:
1) \[-3/2R_1 + R_2\]
2) \[-10R_2 + R_3\]
3) \[1/2R_1\]
4) \[2R_2\]
5) \[-3/2R_2 + R_1\]

Any vector with below relationship can be used with \[\lambda = 1\]
$$a \begin{bmatrix} 1 \\
0 \\
0
\end{bmatrix}$$


```{r}
orig_mat <- matrix(c(1,0,0,2,4,0,3,5,6),nrow=3,ncol=3)
eig_one <- matrix(c(0,0,0,2,3,0,3,5,5),nrow=3,ncol=3)
pracma::rref(eig_one)
print(orig_mat %*% matrix(c(1,0,0)))
print(1*matrix(c(1,0,0)))
```
\[\epsilon_{\lambda=6}\]
$$X = \begin{bmatrix}1-6 & 2  & 3 \\
0 & 4 - 6 & 5 \\
0 & 0 & 6 - 6
\end{bmatrix}$$

Row Reduction:
1) \[R_2 + R_1\]
2) \[-1/5R_1\]
3) \[-1/2R_2\]

Any vector with below relationship can be used with \[\lambda = 6\]
$$ a \begin{bmatrix} 1.6x_3 \\
2.5x_3\\
1
\end{bmatrix}$$

```{r}
eig_two <- matrix(c(-5,0,0,2,-2,0,3,5,0),nrow=3,ncol=3)
pracma::rref(eig_two)
print(orig_mat %*% matrix(c(1.6,2.5,1)))
print(6*matrix(c(1.6,2.5,1)))
```


\[\epsilon_{\lambda=4}\]
$$X = \begin{bmatrix}1-4 & 2  & 3 \\
0 & 4 - 4 & 5 \\
0 & 0 & 6 - 4
\end{bmatrix}$$

Row Reduction:
1) \[1/2R_3\]
2) \[1/5R_2\]
3) \[-1/3R_1\]
4) \[R_2+R_1\]
5) \[-1R_2+R_3\]

Any vector with below relationship can be used with \[\lambda = 4\]
$$ a \begin{bmatrix} 2/3x_2 \\
1\\
0
\end{bmatrix}$$

```{r}
eig_three <- matrix(c(-3,0,0,2,0,0,3,5,2),nrow=3,ncol=3)
pracma::rref(eig_three)
print(orig_mat %*% matrix(c(2,3,0)))
print(4*matrix(c(2,3,0)))
```

